I don't know, are you sure we want to do this?\\ Sea level adaptation decisions under uncertainty

### 2016-12-31 submission/paperMD2.tex
final edits on submission

\begin{abstract}
Sea level rise has serious consequences for harbor infrastructure, storm drains and sewer systems, and many other issues. Adapting to sea level rise requires comparing different possible adaptation strategies, comparing the cost of different actions (including no action), and assessing where and at what point in time the chosen strategy should be implemented. All these decisions must be made under considerable uncertainty--in the amount of sea level rise, in the cost and prioritization of adaptation actions, and in the implications of no action. Here we develop two illustrative examples: for Bergen on Norway's west coast and for Esbjerg on the west coast of Denmark, to highlight how technical efforts to understand and quantify uncertainties in hydrologic projections can be coupled with concrete decision-problems framed by the needs of the end-users using statistical formulations. Different components of uncertainty are visualized. We demonstrate the value of uncertainties and show for example that failing to take uncertainty into account can  result in the median projected damage costs being an order of magnitude smaller. 
\end{abstract}

@@

Esbjerg, on the southwest coast of Jutland, is the fifth-largest city in Denmark and the largest urban area in the region. The city hosts one of the largest harbors  in Denmark, which serves as a focal point for offshore activities in the North Sea, including the continued development of offshore wind power and extensive activities related to the extraction of oil and gas. As a result critical infrastructures and commercial buildings figure prominently in the coastal zone. Esbjerg is frequently subject to substantial storms and storm surges, causing severe flooding of the harbor and the city. The highest since records began in 1874 was recorded in 1981, where the harbor was completely flooded and the water level reached  433 cm above the norm, causing massive economic losses.  More generally, storm surges causing water levels in Esbjerg to rise to between 2 and 3 meters have quadrupled over the last four decades according to local records, whereas half of the most severe events have taken place since 1975.

@@

The remainder of the paper is organized as follows. Section \ref{sealevelproj} describes our approach to projecting sea level. In sections \ref{decision_tools_PartI} and \ref{decision_tools_PartII} we describe the type of decision problems that we are going to attack. We apply these tools to sea level projections for Bergen, Norway, and for Esbjerg, Denmark in section \ref{cases}. In section \ref{unc} we demonstrate the consequences of ignoring the uncertainty in the projections, and the paper is closed with conclusions in section \ref{disc}.

@@

\subsection{Local sea level}
In order to get from global sea level projections to local ones, it is important to note that sea level rise is not uniform over the globe. Glacial and land ice melting affect the local sea level differently depending on where the melted ice is located.
%Gravitational effects
Another major effect in Fennoskandia is the land rise due to isostatic rebound from the glaciers of the last ice age. 
Again, we will use historical data to relate global sea level to isostatically corrected local sea level using a time series regression model. 
The local sea level projections are then obtained by first relating projected temperature to global sea level, and then relating the global sea level to the local one. Each climate model temperature projection yields a different local sea level projection. The local sea level projection based on the i'th climate model for years beyond 2000 is estimated as
\begin{linenomath*}
\[H_t^{loc,i} + \gamma (t -2000 ) = {\hat b} H_t^{gl,i}  + {\varepsilon _t},\]
\end{linenomath*}
where $\gamma$ is the annual land rise rate, $t$ denotes year,  $ {\hat b} $ is the regression coefficient relating global to local sea level and the $\varepsilon _t$ are Gaussian errors.

@@

Climate change adaptation is currently moving from theory to practice with practitioners needing to decide how to begin adapting. This implies an increasing need for detailed economic analysis and appraisal of options   \citep{Downing2012}. Robust decision-making approaches can incorporate various types of uncertainty, including a broad range of climate scenarios \citep{Dittrich&2016}. These methods can be classified as science-first or policy-first approaches. The former type has a ``predict-then-act'' foundation, which starts with climate projections and impact assessments, not linked to any specific adaptation choices \citep{Jones&2014}. The latter starts out with the formulated adaptation plans, and their functioning is tested against different future projections \citep{Dittrich&2016}.

Here, we take a policy-first approach where we test a current adaptation plan against different sea level and damage projections and the inclusion of different sources of uncertainty. We focus on what this implies for the timing of adaptation measures and the implications of including uncertainty. In particular, we employ a probabilistic extension of the framework described by \citep{Fankhauser&1999} in which we obtain a probabilistic distribution for the net present value damage in a given year for various adaptation options. The probabilistic distribution is constructed by considering uncertainty in the local sea level projections, in the annual damage costs, and in the effect of changes in sea level on the annual damage costs.

@@

\subsection{Limitations of the decision framework}
The main limitation of this light touch decision framework is that we have significantly simplified the assessment of the effect of sea level rise on the damage costs. In particular, the linear extrapolation of the results reported in \citet{Hallegatte&2013} might provide a conservative estimate of the effect of extreme sea level rise. However, with only two data points, extrapolation approaches such as a power law or exponential growth seem difficult to justify.

Alternatively, a modeling framework similar to that of \citet{Hallegatte&2013} could be applied directly to a larger range of potential changes in sea level. The elements of such a framework might include an appropriate social discount rate, valuing environmental goods in monetary terms, incorporate socio-economic assumptions and long-term policy goals of decision makers, as well as that climate change is often not the only driver that decision makers should consider, therefore costs and benefits should be studied in a wider context \citep{Dittrich&2016}.

@@

\section{Risk mapping}
\label{decision_tools_PartII}

The initial scoping of climate adaptation in Esbjerg \citep{Esbjerg2014} has been informed by a limited set of floods maps representing different scenarios corresponding, respectively, to flooding caused by storm surges, heavy rainfall and rising ground water levels. The flood maps themselves are produced using simplified modelling approaches representing present day climate conditions, i.e. they do not consider the expected changes in e.g. sea level rise and rainfall characteristics caused by climate change. They also do include an explicit representation of the urban drainage system. Thus, in the context of making decisions on adaptation this preliminary mapping is primarily meant as a tool for identifying high-risk areas, e.g. in terms of avoiding future urban development into such areas, and as a precurser for much more detailed (and resource intensive) local hydrological and economical modelling efforts, e.g. to pave the way for deciding upon cost-effective adaptation measures in the second phase of the plan. As a result what is of relevance to the stakeholders at this stage is not the mapping of hydrological hazards by itself, but rather the mapping of risk, which compounds hazard with (economic) valuation of its consequences for any given map area or pixel. In general, this may be expressed as the probability, e.g. of a certain flood depth, derived from the urban flood model times a damage function associating the flood depth with the chosen measure of cost (e.g. \citet{Halsnaes&2015}). The damage function - as in the case of Bergen - is often expressed in economic terms, however, in Esbjerg a categorical approach has been preferred, wherein a value between 1 and 10 is allocated to each map object, cf. Table \ref{tab:valuemap}.

\begin{table}
\begin{center}
\begin{tabular}[]{l | l | c | c }

Map object&Description&Unit&Value\\
\hline
Important buildings & E.g. industry, hospitals and other public buildings & m2 & 10\\
Other buildings & All other building types than the above & m2 & 8\\
Critical infrastructure & Waste water treatment plant, railroads, etc. & m2 & 10\\
Roads & Essentially all types of roads from main roads to & m2 & 6\\
& minor paths & & \\
Cultural heritage & Including cemetaries, historical landscape & m2 & 0.1-4\\
Natural systems & Included protected areas and sports facilities & m2 & 0.1\\
\hline
\end{tabular}
\end{center}
\caption{Examples of values allocated to different map objects in Esbjerg, prescribed by the end-users \citep{Esbjerg2014}.}
\label{tab:valuemap}
\end{table}

@@

Annual damage costs for the Bergen case study are obtained from the Norwegian Natural Perils Pool (NPP;  data are available at \url{https://www.finansnorge.no/statistikk/skadeforsikring/Naturskadestatistikk-NASK/}). The NPP data are available for the period 1980-2015 and are aggregated to a county level. For improved parameter estimation, we include the data from Rogaland county which is the county directly south of Hordaland and with similar characteristics. We use a discount rate of 4\% for the first 40 years of the analysis, a rate of 3\% for 40 to 75 years into the future and a rate of 2\% beyond 75 years (cf. Section 5.8 of \citet{DiscountRate}).

Storm surge data for Esbjerg ate obtained from the Danish Coastal Authority \citep{sealevel2012}.

@@

Figure~\ref{fig:BergenDamageDist} shows the histogram of observed annual damage costs for Bergen and the associated Burr distribution. The parameter estimates are $\hat{\alpha} = 7.84$ (3.6), $\hat{\gamma} = 0.40$ (0.04) and $\hat{\theta} = 0.007$ (0.01). \citet{bergenreport} discuss several different adaptation options for Bergen. In Figure~\ref{fig:TotalDamageBergen} we consider the optimal timing of an adaptation option that includes two inner barriers at V{\aa}gen and Damg{\aa}rdssundet, that is, one on each side of central Bergen. The combined construction cost of the two barriers for a protection against 75 cm sea level rise is estimated at 1.13 billion NOK (2015 level)\footnote{100 NOK is about 11 EUR.}. As the inner barriers only provide partial areal protection, we assume that they have a mitigating effect on 50\% of damages incurred by storm surges.

\citet{bergenreport} conclude that adaptation measures need not be implemented before 2065. Applying the methodology from section \ref{decision_tools_PartI} we find that the optimal time of building the barriers is almost two decades earlier, in 2047 (Figure \ref{fig:TotalDamageBergen}), and that by the year 2100 this decision will on average save about 1/3 of the median damage costs without adaptation (Figure \ref{fig:NoAction}). At the high end (or the 95th percentile of the damage distribution) adaptation saves even more, about 43\%.

Furthermore, \citet{bergenreport} consider a far-reaching protection measure consisting of an outer barrier that would protect the whole metropolitan area. The cost of this construction is estimated to be at least 34 billion NOK (2015 level). It is assessed that such a construction is not viable due to large environmental and economic consequences. Our estimates of the total damage (Figure 7) support this conclusion.

@@

\begin{figure}[!hbpt]
\begin{center}
\includegraphics[width=\linewidth]{FloodMaps_Harbour_RP20_RP100_dif2.png}
\caption{Flood extents and depths for the city of Esbjerg in year 2100 during storm surges with a return period of 20-years (RP20) and 100-years (RP100) with (a) no sea level rise and with sea level rise corresponding to RCP 85 (5th percentile (b), median (c) and 95th percentile (d)). }
\label{fig:esbjerg}
\end{center}
\end{figure}

@@

In many cases sea level rise projections are given as a single number for each scenario, usually the mean or median of the ensemble of projections from different climate models (e.g. \citet{climateimpactgroup}). Sometimes the spread of the ensemble is used to assess the uncertainty in the projections (e.g., the Norwegian Environmental Agency recommends using the upper ensemble value for RCP 8.5 as the basis for planning decision, pers. comm. from Even Nilsson, Norwegian Mapping Authority). In our analysis there are two more sources of uncertainty, namely the two regression models. Figure \ref{fig:unc} shows the single number (vertical black line), the ensemble spread (histogram), the uncertainty including only the global model (red) and the full uncertainty (blue) for Bergen and Esbjerg projections of sea level rise relative to 1999 under RCP 8.5. For Bergen, we see that the ensemble range is about 16 cm, whereas the overall uncertainty range is about 40 cm. While the values for Esbjerg are somewhat higher, there is slightly less uncertainty in the projections.

\begin{figure}[!hbpt]
\begin{center}
  \includegraphics[width=0.45\linewidth]{unc.pdf}
  \includegraphics[width=0.45\linewidth]{ExbjergUncertainty.pdf}
\caption{2050 Bergen (left) and Esbjerg (right) sea level projections with uncertainty due to different sources for RCP 8.5. The black vertical line is the median projection (with no uncertainty), while the gray histogram corresponds to the spread of the climate models, the red curve adds the uncertainty due to the relation between global temperature and global sea level, and the blue line that due to downscaling global sea level to Bergen. } 
\label{fig:unc}
\end{center}
\end{figure}

\subsection{Damage costs}
A simplistic analysis of projected total damage costs for the year 2100, not taking into account the uncertainty, would use  the median  historical damage cost multiplied by  the median damage effect factor at 2100 at the median sea level rise projected for 2100. Performing this exercise for every year 2016-2100 yields a total (discounted) damage cost of 338 million NOK (the grey vertical line in Figure \ref{fig:costunc}). Similar results are obtained when allowing sea level or effect factor to vary, holding the other quantities at the median (yellow and purple dots on top of Figure \ref{fig:costunc}). However, allowing only the damage cost to vary yields a median total cost of 3.85 billion (green dot on top of Figure \ref{fig:costunc}). The appropriate uncertainty analysis for our model should draw each of sea level, effect factor and damage cost at random from their distributions for 2016-2100. This corresponds to a total median cost of 3.15 billion NOK, over 9 times higher than the simplistic value. Over 99\% of the costs in our simulation are higher than the simplistic median.

\begin{figure}[!hbpt]
\begin{center}
\includegraphics[width=\linewidth]{UncertaintyLog.pdf}
\caption{Simulated distribution of total cumulated damage cost for 2016-2100 without adaptation on log-log scale (black histogram). We also show the distributions of costs varying only one aspect of the uncertainty (sea level rise in purple, effect multiplier in yellow, and damage cost in green), holding the other two at their median values. The grey vertical line is the result of holding all three factors at their median value. The median of each distribution is shown as a dot on top of the figure.} 
\label{fig:costunc}
\end{center}
\end{figure}

@@

Seen from the perspective of a decision-maker, however, adaption actions are likely to be related to an assessment of risk and not hazard. In this study we have not compounded our simple flood risk estimates with the categorical valuation scheme used by Esbjerg in the first phase of their adaptation plan to produce a risk mapping. Hence arguably the design of the original scheme does not discriminate sufficiently between assets to allow for an objective prioritization based on the underlying data, i.e. buildings are generally all assigned a high value, whereas all other assets are assigned significantly lower value. Using a more elaborate valuation model (such a model is currently being developed by Esbjerg municipality in collaboration with consultants) the uncertainties introduced by the valuation model should also be introduced into the probabilistic modelling chain, since the uncertainties introduced by the underlying economic assumptions and modelling \citep{Halsnaes&2015} may have significant implications on the final results of the risk mapping. One example of this, as already highlighted in the current scheme (Table \ref{tab:valuemap}), is the valuation of natural areas, which in the context of the Esbjerg case in particularly relates to the marshes close to the nearby city of Ribe that are protected under the code of Natura 2000\footnote{See \url{http://ec.europa.eu/environment/nature/natura2000/index_en.htm}.}. From interviews with the municipality it is evident that assigning an objective value to this area, whether by economic cost or in terms of a score, which aligns with the overall objectives of the adaptation plan is a difficult task, which must be taken into account when making the risk assessment including quantification of the uncertainties operational from the end-users' perspectives.

\section{Conclusions}
\label{disc}
Sea level rise has serious consequences for infrastructure in coastal areas. Our case studies demonstrate that it is possible to take uncertainty into account in deciding when and where to implement adaptation measures, even if one uses a light touch decision-making approaches. Failing to account for uncertainty can result in bad scenarios, such as 95th percentiles, to be an order of magnitude worse than what the planners are expecting. For climate change adaptation in general, it is likely worthwhile to be pessimistic in the planning and in the projections.

We consider our case studies proofs of concept and see these as first steps in a sequence of interactions with local planners and other end-users.The adaptation planning process has to be an iterative and interactive process, as the decision framework provides actionable information to decision-makers, who will then make their own decisions. These decisions can then be incorporated into the current adaptation strategy. Further simulation studies allow a continued loop to identify potential vulnerabilities of the approaches across a wide range of possible futures.

Down the line we plan to develop a flexible and easy-to-use tool kit for decision-making under uncertainty regarding sea level rise. An initial step in this direction is the software used in this analysis, which is publicly available and uses free software.

>>>>

\begin{abstract}
Sea level rise has serious consequences for harbor infrastructure, storm drains and sewer systems, and many other issues. Adapting to sea level rise requires comparing different possible adaptation strategies, comparing the cost of different actions (including no action), and assessing where and at what point in time the chosen strategy should be implemented. All these decisions must be made under considerable uncertainty--in the amount of sea level rise, in the cost and prioritization of adaptation actions, and in the implications of no action. Here we develop two illustrative examples: for Bergen on Norway's west coast and for Esbjerg on the west coast of Denmark, to highlight how technical efforts to understand and quantify uncertainties in hydrologic projections can be coupled with concrete decision-problems framed by the needs of the end-users using statistical formulations. Different components of uncertainty are visualized. We demonstate the value of uncertainties and show for example that failing to take uncertainty into account can  result in the median projected damage costs being an order of magnitude smaller. 
\end{abstract}

@@

Esbjerg, on the southwest coast of Jutland, is the fifth-largest city in Denmark and the largest urban area in the region. The city hosts one of the largest harbors  in Denmark, which serves as a focal point for offshore activities in the North Sea, including the continued development of offshore wind power and extensive activities related to the extraction of oil and gas. As a result critical infrastructures and commercial buildings figure prominently in the coastal zone. Esbjerg is frequently subject to substantial storms and storm surges, causing severe flooding of the harbor and the city. The highest since records began in 1874 was recorded in 1981, where the harbor was completely flooded and the water level reached  433 cm above the norm, causing massive economic losses.  More generally, storm surges causing water levels in Esbjerg to rise to between 2 and 3 metres have quadrupled over the last four decades according to local records, whereas half of the most severe events have taken place since 1975.

@@

The remainder of the paper is organized as follows. Section \ref{sealevelproj} describes our approach to projecting sea level. In sections \ref{decision_tools_PartI} and \ref{decision_tools_PartII} we describe the type of decision problems that we are going to attack. We apply these tools to sea level projections for Bergen, Norway, and for Esbjerg, Denmark in section \ref{cases}. In section \ref{unc} we demonstrate the consequences of ignoring the uncertainty in the projections, and the paper is closed with a summary and discussion in section \ref{disc}.

@@

\subsection{Local sea level}
In order to get from global sea level projections to local ones, it is important to note that sea level rise is not uniform over the globe. Glacial and land ice melting affect the local sea level differently depending on where the melted ice is located.
%Gravitational effects
Another major effect in Fennoskandia is the land rise due to isostatic rebound from the glaciers of the last ice age. 
Again, we will use historical data to relate global sea level to isostatically corrected local sea level using a time series regression model. 
The local sea level projections are then obtained by first relating projected temperature to global sea level, and then relating the global sea level to the local one. Each climate model temperature projection yields a different local sea level projection. The local sea level projection based on the i'th climate model for years beyond 2000 is estimated as
\begin{linenomath*}
\[H_t^{loc,i} + \gamma (t -2000 ) = {\hat b} H_t^{gl,i}  + {\varepsilon _t},\]
\end{linenomath*}
where $\gamma$ is the annual land rise rate, $t$ denotes year,  $ {\hat b} $ is the regression coefficient relating global to local sea level and the $\varepsilon _t$ are Gaussian errors..

@@

There is an increasing need for more detailed economic analysis, including simple methods and tools for assessment of options, especially since as Downing (2012) recognizes, adaptation is moving from theory to practice, and practitioners try to deal with how to begin adapting. This leads to an increasing need and interest in the appraisal of options.

Several economic decision support tools and methods exist for adaptation assessment under uncertainty. Robust decision-making approaches are able to better incorporate uncertainty and a broad range of climate scenarios to capture as much of the uncertainty on future climates as possible (Dittrich et al. 2016). These approaches can be classified according to a science-first or policy-first approach. The former has a ``predict-then-act'' foundation, which starts with climate projections and impact assessments, not linked to any specific adaptation choices (Jones et al., 2014). The latter starts out with the formulated adaptation plans and not impacts, and their functioning is tested against different future projections (Dittrich et al. 2016).

We take a policy-first approach, in which we test a current adaptation plan against different sea level and damage projections and the inclusion of different sources of uncertainty. We focus on what this implies for the timing of adaptation measures and the implications of including uncertainty. In particular, we employ a probabilistic extension of the framework described by \citep{Fankhauser&1999} in which we obtain a probabilistic distribution for the net present value damage in a given year for various adaptation options. The probabilistic distribution is constructed by considering uncertainty in the local sea level projections, in the annual damage costs, and in the effect of changes in sea level on the annual damage costs.

@@

\subsection{Limitations of the decision framework}
The main limitation of this light touch decision framework is that we have significantly simplified the assessment of the effect of sea level rise on the damage costs. In particular, the linear extrapolation of the results reported in \citet{Hallegatte&2013} might provide a conservative estimate of the effect of extreme sea level rise. However, with only two data points, extrapolation approaches such as a power law or exponential growth seem difficult to justify\`.

Alternatively, a modeling framework similar to that of \citet{Hallegatte&2013} could be applied directly to a larger range of potential changes in sea level. The elements of such a framework might include an appropriate social discount rate, valuing environmental goods in monetary terms, incorporate socio-economic assumptions and long-term policy goals of decision makers, as well as that climate change is often not the only driver that decision makers should consider, therefore costs and benefits should be studied in a wider context (Dittrich et al, 2016).

@@

\section{Risk mapping}
\label{decision_tools_PartII}
The initial scoping of climate adaptation in Esbjerg {\color{blue}(cite: Esbjerg2014)} has been informed by a limited set of floods maps representing different scenarios corresponding, respectively, to flooding caused by storm surges, heavy rainfall and rising ground water levels. The flood maps themselves are produced using simplified modelling approaches representing present day climate conditions, i.e. they do not consider the expected changes in e.g. sea level rise and rainfall characteristics caused by climate change. They also do include an explicit representation of the urban drainage system. Thus, in the context of making decisions on adaptation this preliminary mapping is primarily meant as a tool for identifying high-risk areas, e.g. in terms of avoiding future urban development into such areas, and as a precurser for much more detailed (and resource intensive) local hydrological and economical modelling efforts, e.g. to pave the way for deciding upon cost-effective adaptation measures in the second phase of the plan. As a result what is of relevance to the stakeholders at this stage is not the mapping of hydrological hazards by itself, but rather the mapping of risk, which compounds hazard with (economic) valuation of its consequences for any given map area or pixel. In general, this may be expressed as the probability, e.g. of a certain flood depth, derived from the urban flood model times a damage function associating the flood depth with the chosen measure of cost (e.g. {\color{blue}(cite:Halsnaes2015)}. The damage function - as in the case of Bergen - is often expressed in economic terms, however, in Esbjerg a categorical approach has been preferred, wherein a value between 1 and 10 is allocated to each map object, c.f. Table \ref{tab:valuemap}

\begin{table}
\begin{center}
\begin{tabular}[]{c | c | c | c }

Map object&Description&Unit&Value\\
\hline
Important buildings&E.g. industry, hospitals and other public buildings&m2&10\\
Other buildings&All other building types than the above&m2&8\\
Critical infrastructure&Waste water treatment plant, railroads, etc.&m2&10\\
Roads&Covers essentially all types of roads from main roads to minor paths&m2&6\\
Cultural heritage&Including cemetaries, historical landscape&m2&0.1-4\\
Natural systems&Included protected areas and sports facilities&m2&0.1\\
\hline
\end{tabular}
\end{center}
\caption{Examples of values allocated to different map objects in Esbjerg, prescribed by the end-users {\color{blue}(cite: Esbjerg2014)}.}
\label{tab:valuemap}
\end{table}

@@

Annual damage costs for the Bergen case study are obtained from the Norwegian Natural Perils Pool (NPP;  data are available at \url{https://www.finansnorge.no/statistikk/skadeforsikring/Naturskadestatistikk-NASK/}). The NPP data are available for the period 1980-2015 and are aggregated to a county level. For improved parameter estimation, we include the data from Rogaland county which is the county directly south of Hordaland and with similar characteristics. We use a dicsount rate of 4\% for the first 40 years of the analysis, a rate of 3\% for 40 to 75 years into the future and a rate of 2\% beyond 75 years (cf. Section 5.8 of \citet{DiscountRate}).

Storm surge data for Esbjerg are obtained from the Danish Coastal Authority \citep{sealevel2012}.

@@

Figure~\ref{fig:BergenDamageDist} shows the histogram of observed annual damage costs for Bergen and the associated Burr distribution. The parameter estimates are $\hat{\alpha} = 7.84$ (3.6), $\hat{\gamma} = 0.40$ (0.04) and $\hat{\theta} = 0.007$ (0.01). \citet{bergenreport} discuss several different adaptation options for Bergen. In Figure~\ref{fig:TotalDamageBergen} we consider the optimal timing of an adaptation option that includes two inner barriers at V{\aa}gen and Damg{\aa}rdssundet, that is, one on each side of central Bergen. The combined construction cost of the two barriers for a protection against 75 cm sea level rise is estimated at 1.13 billion NOK (2015 level)\footnote{100 NOK is about 11 EUR.}.

Applying the methodology from section \ref{decision_tools_PartI} we find that the optimal time of building the barriers is in 2046 (Figure \ref{fig:TotalDamageBergen}), and that by the year 2100 this decision will on average save about 1/3 of the median damage costs without adaptation (Figure \ref{fig:NoAction}). At the high end (or the 95th percentile of the damage distribution) adaptation saves even more, about 43\%.

@@

\begin{figure}[!hbpt]
\begin{center}
\includegraphics[width=\linewidth]{Floodmaps_Harbour.png}
\caption{Flood extents and depths for the city of Esbjerg in year 2100 during storm surges with a 20 year return period (upper row) and a 100 year return period (lower row) with (a) no sea level rise and with sea level rise corresponding to RCP 85 (5th percentile (b), median (c) and 95th percentile (d)). }
\label{fig:esbjerg}
\end{center}
\end{figure}

@@

In many cases sea level rise projections are given as a single number for each scenario, usually the mean or median of the ensemble of projections from different climate models (e.g. \citet{climateimpactgroup}). Sometimes the spread of the ensemble is used to assess the uncertainty in the projections (e.g., the Norwegian Environmental Agency recommends using the upper ensemble value for RCP 8.5 as the basis for planning decision, pers. comm. from Even Nilsson, Norwegian Mapping Authority). In our analysis there are two more sources of uncertainty, namely the two regression models. Figure \ref{fig:unc} shows the single number (vertical black line), the ensemble spread (histogram), the uncertainty including only the global model (red) and the full uncertainty (blue) for Bergen projections of sea level rise relative to 1999 under RCP 8.5. We see that the ensemble range is about 16 cm, whereas the overall uncertainty range is about 40 cm.

\begin{figure}[!hbpt]
\begin{center}
\includegraphics[width=0.5\linewidth]{unc.pdf}
\caption{2050 Bergen sea level projections with uncertainty due to different sources for RCP 8.5. The black vertical line is the median projection (with no uncertainty), while the gray histogram corresponds to the spread of the climate models, the red curve adds the uncertainty due to the relation between global temperature and global sea level, and the blue line that due to downscaling global sea level to Bergen. } 
\label{fig:unc}
\end{center}
\end{figure}

\subsection{Damage costs}
A simplistic analysis of projected damage costs at the year 2100, not taking into account the uncertainty, would use  the median  historical damage cost multiplied by  the median damage effect factor at 2100 at the median sea level rise projected for 2100. This yields a damage cost of NOK 338 million (the grey vertical line in Figure \ref{fig:costunc}). Similar results obtain when allowing sea level or effect factor to vary, holding the other quantities at the median (yellow and purple dots on top of Figure \ref{fig:costunc}). However, allowing only the damage cost to vary yields a median cost of 3.85 billion (green dot on top of Figure \ref{fig:costunc}). The appropriate uncertainty analysis for our model should draw each of sea level, effect factor and damage cost at random from their 2100 distributions. This corresponds to a median cost of 3.15 billion NOK, over 9 times higher than the simplistic value. Over 99\% of the costs in our simulation are higher than the simplistic median.

\begin{figure}[!hbpt]
\begin{center}
\includegraphics[width=0.8\linewidth]{UncertaintyLog.pdf}
\caption{Simulated distribution of total damage cost for 2100 without adaptation on log-log scale (black histogram). We also show the distributions of costs varying only one aspect of the uncertainty (sea level rise in blue, effect multiplier in yellow, and damage cost in green), holding the other two at their median values. The grey vertical line is the result of holding all three factors at their median value. The median of each distribution is shown as a dot on top of the figure.} 
\label{fig:costunc}
\end{center}
\end{figure}

@@

Seen from the perspective of a decision-maker, however, adaption actions are likely to be related to an assessment of risk and not hazard. In this study we have not compounded our simple flood risk estimates with the categorical valuation scheme used by Esbjerg in the first phase of their adaptation plan to produce a risk mapping. Hence arguably the design of the original scheme does not discriminate sufficiently between assets to allow for an objective prioritization based on the underlying data, i.e. buildings are generally all assigned a high value, whereas all other assets are assigned significantly lower value. Using a more elaborate valuation model (such a model is currently being developed by Esbjerg municipality in collaboration with consultants) the uncertainties introduced by the valuation model should also be introduced into the probabilistic modelling chain, since the uncertainties introduced by the underlying economic assumptions and modelling ({\color{blue}(cite:Halsnaes2015)}) may have significant implications on the final results of the risk mapping. One example of this, as already highlighted in the current scheme (Table \ref{tab:valuemap}), is the valuation of natural areas, which in the context of the Esbjerg case in particularly relates to the marshes close to the nearby city of Ribe that are protected under the code of Natura 2000\footnote{See \url{http://ec.europa.eu/environment/nature/natura2000/index_en.htm}.}. From interviews with the municipality it is evident that assigning an objective value to this area, whether by economic cost or in terms of a score, which aligns with the overall objectives of the adaptation plan is a difficult task, which must be taken into account when making the risk assessment including quantification of the uncertainties operational from the end-users' perspectives.

%\subsection{...}

\section{Conclusions and discussion}
\label{disc}
Our case studies demonstrate that it is possible to take uncertainty into account in deciding when and where to implement adaptation measures, even if one uses a light touch decision-making approach. If one fails to do so, bad scenarios, such as 95th percentiles, can be a order of magnitude worse than what the planners are expecting. It is likely worthwhile to be pessimistic in the planning and in the projections.

We consider our case studies proofs of concept, which will be first steps in a sequence of interactions with local planners and other end-users.This has to be an iterative and interactive process, as the decision framework provides actionable information to decision-makers, who will then make their own decisions. These decision can then be incorporated into the current adaptation strategy. Further simulation studies allow a continued loop to identify potential vulnerabilities of the approaches across a wide range of possible futures.

Down the line we plan to develop a flexible and easy-to-use tool kit for decision-making under uncertainty regarding sea level rise. An initial step in this direction is the software used in this analysis, which is publicly available and using free software.
